The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-20 R/old_eda_paper.Rmd
Edit eda section.

---
title: "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth"
authors:
  - name: Dewi Amaliah
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dama0007@student.monash.edu
  - name: Dianne Cook
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: ""
  - name: Nicholas Tierney
    thanks: ""
    department: ""
    affiliation: ""
    location: ""
    email: ""
abstract: > 
 The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data is never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets, and the code to produce them are provided in an accompanying open source R package, called `yowie`. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks, and providing more convenient samples of the data to potential users. 
bibliography: references.bib
biblio-style: unsrt
#preamble: >
header-includes:
  - \usepackage{tcolorbox}
  - \usepackage{fontawesome}
output: 
  #rticles::arxiv_article:
    #keep_tex: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
date: "24/02/2021"
always_allow_html: true
keywords: "Data cleaning; Data tidying; Longitudinal data; NLSY79; Open data; Initial data analysis; Outlier detection; Robust linear regression"
---

```{r setup, echo = FALSE, cache = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, warning = FALSE, 
               message = FALSE,
               cache = TRUE, 
               cache.path = "paper_cache/",
               fig.retina = 2,
               fig.path = "figures/",
               comment = "#>",
               fig.align = 'center')

read_chunk(here::here("paper/scripts/data-preprocessing.R"))
read_chunk(here::here("paper/scripts/ida.R"))
read_chunk(here::here("paper/scripts/eda.R"))
```

```{r load-pkgs}
```

# Introduction {#intro}

"Open data" is data that is freely accessible, modifiable, and shareable by anyone for any purpose [@opendata]. This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call "wild data" because it requires substantial cleaning and tidying to tame it into textbook shape. @HuebnerMariannePhD2016Asat emphasize that making the data cleaning process accountable and transparent is imperative. Documenting the data cleaning is essential for the integrity of downstream statistical analyses and model building [@HuebnerMarianne2020Haar]. 
Data cleaning can be considered to be a part of what is called "initial data analysis (IDA)" [@Chatfield1985TIEo]. In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis, which @tukey defines as detective work to learn from data, either numerically or graphically, before confirmatory data analysis is performed, which encompasses IDA. @DasuTamraparni2003Edma say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80% of the data mining task.

Despite its importance, this IDA stage is often undervalued and neglected [@Chatfield1985TIEo]. There are few research papers that document the data cleaning [@WickhamHadley2014TD]. Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties [@HuebnerMarianne2020Haar].

This paper aims to demonstrate the steps of IDA, tidying, cleaning, and documenting the process, for a prominent open data source, National Longitudinal Survey of Youth (NLSY) 1979, henceforth referred to NLSY79. This data has been playing an important role in research in various disciplines, including but are not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN]. In addition, The NLSY is considered a survey with high retention rates and carefully designed, making it suitable for a life course research [@MichaelRPergamit2001DWTN; @eliznlsy]. According to @eliznlsy, thousand of articles, and hundreds of book chapters and monographs has incorporated the NLSY data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the NLSY79 data sets [@MichaelRPergamit2001DWTN].

@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling. Our aim is to refresh this textbook data to be more up to date. Here, we investigate the process of getting from the raw NLSY79 to this textbook data set. However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.

This paper is structured to have 5 sections. Section \@ref(database) describes the original data source. Section \@ref(cleaning) presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and initial data analysis to find and treat the anomalies in the data. We also gave examples of exploratory data analysis using the clean data in Section \@ref(eda). Finally, Section \@ref(summary) summarises the paper.

# The NLSY79 {#database}

## Database

The NLSY79 is a longitudinal survey held by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth and born between 1957-1964 [@nlsy79]. The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979. It was comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military. In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects are the 1,079 members of the military sample and 1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively. Hence, 9,964 respondents remain in the eligible samples. The surveys were conducted annually from 1979 to 1994 and biennially thereafter. Data are now available from Round 1 (1979 survey year) to Round 28 (2018 survey year).

[FILL] We should include how many variables there are in the data so people get a grasp of the extensive coverage of the variables.

Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education; training and achievement; household, geography, and contextual variables; dating, marriage, cohabitation; sexual activity, pregnancy, and fertility; children; income, assets and program participation; health; attitudes and expectations; and crime and substance use.

There are two ways to conduct the interview of the NLSY79, which are face-to-face interviews or by telephone. In recent survey years, more than 90 percent of respondents were interviewed by telephone [@eliznlsy].

## Target Data

As mentioned in Section \@ref(intro), this paper aims to refresh the NLSY79 data as used in Singer and Willet's textbook (@SingerJudithD2003Alda). This data set contains the information of yearly mean hourly wages from the NLSY79 cohort, with education and race as covariates. This data set represents the measurement of male high-school dropouts, aged from 14 to 17 years old when first time measured. Accordingly, to get this data, we have two additional target data, NLSY79 cohort's demographic profiles and wages data of people who completed education up to high school.

Since there is no information on high school dropouts in Singer and Willet's book, we define high school dropouts as respondents who only completed either $9^{th}$, $10^{th}$, $11^{th}$ grade, or who completed $12^{th}$ when their age is at least 19 years old (means that these IDs being dropped out in a certain year, but they returned back to high school to complete it).

# Data Cleaning {#cleaning}

In this section, we outline the steps taken to the download the raw data from the database provided at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} [@nlsy79]. Specifically, we only extract a subset of the variables in the database as shown in \ref{box:data}.

## Getting and Tidying the Data

The NLSY79 data are stored in a [database](https://www.nlsinfo.org/content/cohorts/nlsy79/get-data) and could be downloaded by variables. Several variables are available for download and could be browsed by index. For the wages data set, we only extracted these variables:

::: {.tcolorbox data-latex="[label=\"box:data\", title=Getting the data]" }
\faDatabase National Longitudinal Survey of Youth 1979

\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCloudDownload] Downloaded all of variables in Highest grade completed.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job $\triangleright$ Hours worked and Hourly wages
\begin{itemize}
\item[\faCloudDownload] Downloaded all of the variables in Hours worked.
\item[\faCloudDownload] Downloaded all of the variables in Hourly wages
  Both hours worked and hourly wages are recorded by the job, up to five jobs for each id/subject. 
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics 
\begin{itemize}
\item[\faCloudDownload] Downloaded year and month of birth, race, and sex variables. There are two versions of the year and month of birth, i.e., 1979 and 1981 data. We downloaded these two versions.
\end{itemize}
\end{itemize}
\end{itemize}

:::

The downloaded data set came in a csv (NLSY79.csv) and dat (NLSY79.dat) format. We only used the .dat format. It also came along with these files:

- NLSY79.NLSY79: Tagset of variables that can be uploaded to the website to recreate the data set.
- NLSY79.R: R script provided automatically by the database for reading the data into R and converting the variables' names and label into something more sensible. We utilized this code to do an initial data tidying. It produced two data set,  `categories_qnames` (the observations are stored in categorical/interval values) and `new_data_qnames` (the observations are stored in integer form).

```{r raw-data}
```

According to @WickhamHadley2014TD, tidy data sets comply with three rules. Firstly, each variable forms a column. Secondly, each observation forms a row. Lastly, each type of observational unit forms a table. Unfortunately, the `new_data_qnames` did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used `tidyr` [@tidyr], `dplyr` [@dplyr], and `stringr` [@stringr] to do this job.

```{r untidy-data}
```

### Tidying demographic variables

`Age in 1979`, `gender`, `race`, `highest grade completed`, and the `year when the highest grade completed` are the variables that we want to use in the cleaned data set.

Age in 1979 are derived from year and month of birth variables in the raw data. The variables have two versions, which are the 1979 version and the 1981 version. We only used the 1979 data and did a consistency check of those years and flag the inconsistent observations.

```{r dob-tidy, echo=TRUE}
```

For `gender` and `race`, we only rename the column names. It is worth noting that using these two variable needs special attention. Gender, as reported in the data, only has two categories, which is recognised today as inadequate. Gender is not binary. Further, race is as reported in the database. When doing analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

```{r demog-tidy}
```

The next step is tidying the `highest grade completed` variable. This variable came with several version in each year. We choose the revised May data because it seemed to have less missing and presumably has been checked. However, there is no revised May data for 2012, 2014, 2016, and 2018. Thus, we just use the ordinary May data for these years. The code for tidying this variable is provided in the supplementary materials.

```{r tidy-grade}
```

Further, `highest grade completed` is measured and could be updated in each period of the survey. Hence, we only used the highest grade completed ever and derived the year when it is completed. The code for tidying this variable is also available in the supplementary materials.

```{r tidy-hgc}
```

Finally, we get all of the demographic profiles of the NLSY79 cohort. We then save this data as `demog_nlsy79`.

```{r full-demog}
```

### Tidying employment variables

The employment data comprises of three variables, i.e., `total hours of work per week`, `number of jobs that an individual has`, and `mean hourly wage`.

`hours worked per week` initially has only one version per job, no choice from 1979 to 1987 (QES-52A). From 1988 onward, when we had more options, we chose the variable for total hours, including time spent working from home (QES-52D). However, in 1993, this variable did not have all the five D variables (the first one and the last one were missing), so we used QES-52A variable instead. In addition, 2008 only had jobs 1-4 for the QES-52D variable (whereas the other years had 1-5), so we use these.

```{r tidy-hours, echo=TRUE}
```

The same way was also deployed to tidy the rate of wage by year and by ID. The difference is that the hourly rate has only one version of each year. The hours of work and the hourly rate are then joined to calculate the number of jobs that a respondent has and their mean hourly wage. Some observations have 0 in their hourly rate and work hours, which is considered an invalid value. Thus, these observations are set to be NA. Besides, there are some observations with unusual hours of work. Here, we also censored the observations with > 84 hours of work a week to be NA.

```{r tidy-rate, echo = TRUE}
```

Since our ultimate goal is to calculate the mean hourly wage, the number of jobs is calculated based on the availability of the `rate_per_hour` information. For example, the number of jobs of ID 1, based on `hours_work`, is 2. However, since the information of hourly rate of `job_02` is not available, the number of jobs is considered as 1.

Further, we calculated the mean hourly wage for each ID in each year using a weighted mean with the hours of work as the weight. However, there is a lot of missing value in `hours_work` variable. In that case, we only calculated the mean hourly wage based on the arithmetic/regular mean method. Hence, we created a new variable to flag whether the mean hourly wage is weighted or regular. Additionally, if an ID only had one job, we directly used their hourly wages information and flagged it as arithmetic mean.

```{r tidy-nojob}
```

The employment and demographic variables are then joined. We also filtered the data to only have the cohort who completed the education up to 12th grade and participated at least five rounds in the survey and save it to an object called `wages`.

```{r wages-demog-hs}
```

## Initial Data Analysis

According to @HuebnerMariannePhD2016Asat, Initial Data Analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis. Moreover, @Chatfield1985TIEo argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

In this paper, we conduct an IDA or a preliminary data analysis to assess the consistency of the data with the cohort information that the NLSY provides. In addition, we also aim to find the anomaly in the wages values using this approach. We mainly use graphical summaries to do the IDA using `ggplot2`[@ggplot2] and `brolgar` [@brolgar].

As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979. Hence, we validate whether all of the respondents were in this range. Additionally, the [NLSY](https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction) also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic). To validate this, we used the `demog_nlsy79`, i.e., the data with the survey years 1979 sample. Table \ref{tab:age-table} and Table \ref{tab:gender-race-table} suggest that the demographic data we had is consistent with the sample information in the database.

```{r age-table}
```

```{r gender-race-table}
```

The next step is that explore the mean hourly wage data. In this case, we only explore the wages data of up to $12^{th}$ grade.  Here, we use visualisation techniques to do IDA.

```{r summarytable}
```

```{r sample-plot, echo = FALSE, fig.cap="A glimpse of the refreshed data. Longitudinal profiles of wages from 1979-2018 are shown for a random sample of 36 individuals, divided into twelve subplots. Most individuals here experienced increasing wages over time, although there is considerable fluctuation from one year to another.", fig.height=4, fig.width=5, out.width="100%"}
```

We take 36 samples randomly from the data and plot them, as shown in Figure \ref{fig:sample-plot}. It implies that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11. The plot also implies that the samples have a different pattern of mean hourly wages. Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.

```{r feature-plot, fig.cap = "Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual (id=39) with an unusual wage relative to their years of data (C). Some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.", fig.width=6, fig.height=2}
```

```{r high-wages}
```

```{r high-wages, fig.cap= "6 out of 45 IDs with extremely high mean hourly wage. Most of the IDs only have one point of high wage.", eval=FALSE}
```

The anomalies are also found in the total hours of work, where some observations work for 420 hours a week in total. According to @MichaelRPergamit2001DWTN, one of the flaws of the NLSY79 employment data is that the NLSY79 collects the information of the working hours since the last interview. Thus, it might be challenging for the respondents to track the within-job hours' changes between survey years, especially for the respondents with fluctuated working hours or seasonal jobs. It even has been more challenging since 1994, where the respondents had to recall two years periods. This shortcoming might also contribute to the fluctuation of one's wages data.

### Replacing extreme values

As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data. Robust linear regression yields an estimation robust to the influence of noise or contamination [@KollerManuel2016rARP]. It also aims to detect the contamination by weighting each observation based on how "well-behaved" they are, known as robustness weight. Observations with lower robustness weight are suggested as an outlier by this method [@KollerManuel2016rARP].

Since we work with longitudinal data, we build the model for each ID instead of the overall data. The robust mixed model could be the best model to be employed in this case. However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data. Thus, the model for each ID is built utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively.

We build the model using the `rlm` function from `MASS` package [@mass]. We set the `mean_hourly_wage` and `year` as the dependent and predictor, respectively. Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm]. However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. Moreover, it should be noted that not all the outliers are due to an error. Instead, it might be that one had reasonably increasing or decreasing wages in a particular period.

To minimize the risk of mistakenly identifying an outlier as an "erroneous outlier", we simulate some thresholds and study how they affect the data. We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback's risk because it still captures the sensible spikes in the data. In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording. After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called `is_pred`.

```{r rlm, eval=FALSE}
```

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. It implies that the fluctuation can still be observed in the data after the treatment. However, the large spikes, which are considered "erroneous outliers", are already eliminated from the data. Hence, the model produces a data set with a more reasonable degree of fluctuation.

```{r compare-data}
```

```{r compare-plot, fig.cap="Comparison between the original and the treated mean hourly wage. The orange line portray the original value of mean hourly wage, while the turquoise line display the mean hourly wages value after the extreme values imputed with the robust linear model's prediction value. We can see that some extreme spikes has been reduced by the model.", fig.height=5, fig.width=6, outwidth="100%"}
```

Further, \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around $350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value, transformed to log scale. The plot implies that the skewness is slightly negative. We also see that the three features are overlapped each other. It indicates that some ID's minimum wages are higher than some ID's maximum wages.

```{r fixed-feature-plot, fig.cap = "Re-make of the summary plots of the fully processed data suggest that the data is now in a reasonable state: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual with an unusual wage relative to their years of data (C). ", fig.width=6, fig.height=2}
```

Finally, we save the imputed data and set the appropriate data type for the variables. As our target data is the mean hourly wage of the high-school dropouts, we then subset the high-school graduate data set to have only the male-high school dropout data. We then make these three data sets and their processing documentation publicly available through an R data container package called `yowie`. The complete flow from the raw data to these data set is displayed in Figure \ref{fig:flow-chart}.

```{r save-data}
```

```{r nrow}
```

## Recap

```{r flow-chart, fig.cap="The stages of data filtering from the raw data to get three datasets contained yowie, n means the number of ID, while n\\_obs means the number of observations.", out.width="93%"}
```

@@

# Summary {#summary}

This paper has performed stages to make open data suitable for textbook data or make it ready for research. In the first stage, we showed the steps performed to get the data from the NLSY79 database. Since the data format is untidy, we showed how the data had been tidied. After that, we conducted an initial data analysis to investigate and screen the quality of the data. We found and fixed the anomalous observations in the data set using the robust linear regression model with its predicted values. We also performed an example of exploratory data analysis using the cleaned data set.

This paper has also demonstrated data cleaning documentation by providing all of the codes in performing data tidying and initial data analysis. Thus, this paper provided an opportunity to continuously refresh the textbook data whenever the updated data is published in the NLSY79 database. It could be done by following the documentation of the code that is provided in this paper.

Moreover, the documentation also includes how we generated the robustness weight and how we decided the threshold of anomalous observations. It is also documented along with the flag of whether an observation is an imputed value or not. Accordingly, if somebody wishes to make another decision, it can be done by making a small change in the code provided. Further, this paper is also supplemented by a `shiny`  [@shiny] app as a simulation tool to customize the weight threshold.

Finally, this paper implies that data providers should design a database that is able to produce tidy data sets. A data provider should also check for data anomalies before the data publishing or at least provides a set of rules or threshold values. For example, in this case, is the threshold of reasonable wages. This will greatly support the data users to validate and set the same understanding of which data are considered outliers. Moreover, providing validation rules would facilitate any established data validation tool, such as `validate` [@validate] package. In this case, we cannot use this handy validation package due to the absence of validation rules.

# Acknowledgements

We would like to thank Aarathy Babu for the insight and discussion during the writing of this paper.

The entire analysis is conducted using `R` [@R] in `rstudio` using these packages: `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], `dplyr` [@dplyr], `readr` [@readr], `tidyr` [@tidyr], `stringr` [@stringr], `purrr` [@purrr], `broom` [@broom], `brolgar` [@brolgar], `patchwork` [@patchwork], `kableExtra` [@kableExtra], `MASS` [@mass], `janitor` [@janitor], `DiagrammeR` [@DiagrammeR], `rsvg` [@rsvg], `webshot` [@webshot], `mgcv` [@mgcv], `tsibble` [@tsibble], and `modelr` [@modelr]. The paper are generated using `knitr` [@knitr], `rmarkdown` [@rmarkdown], and `rticles` [@rticles].

# Supplementary Materials

- **Codes** : R script to reproduce data tidying and cleaning are available in this [Github Repository](https://github.com/numbats/yowie/blob/master/data-raw/data_preprocessing.R).

- **R Package `yowie`**:`yowie` is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package could be accessed [here](https://github.com/numbats/yowie).

- **shiny app**: An web interactive `shiny` app to run a simulation to customize the weight threshold. This app could be accessed [here](https://github.com/numbats/summer-wages-refresh/tree/main/app).

# References

>>>>

---
title: "old_eda_paper"
author: "Dewi"
date: "21/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

### 2021-09-20 paper/paper.Rmd
Tidy plots.
Edit eda section.

@@

# Exploratory Data Analysis {#eda}

This section will provide an example of how data sets could be used to do exploratory data analysis as in textbook data. We are interested in examining how the respondents experience changes in their wages based on their demographic characteristics. Thus, we use `wages` data instead of the refreshed `wages_hs_do` as it only has male respondents.

The statistics we use in measuring the change in wages experience is the slope of the wages of each respondent. 
Using `key_slope` function in `brolgar` [@brolgar], we perform linear regression with the log of mean hourly wage as the response variable and year as the predictor. Since we do modeling, it is important to look at the number of observations from each respondent. Figure \ref{fig:slope-nobs-plot} shows that the number of observations for each individual varies. Moreover, each number of observations seems to have an impact on the slope. Extreme values in slope are observed in a few observations, while the range of slope is more stable with more observations. Accordingly, we trim the data to only include respondents with at least 10 observations.

```{r nobs-plot}
```

```{r slope}
```

```{r slope-nobs}
```

```{r slope-nobs-plot, fig.cap = "A) Distribution of observations. The numbers of rounds varies among respondents. Most of respondents participated in 28 rounds of survey. B) Distribution of wages slope by number of observations. The slopes' range are more stable with more observations."}
```

```{r trim-slope}
```

We then create some visualizations of five respondents with the highest slope and five respondents with the lowest slope, henceforth referred to as top-5 slopes and bottom-5 slopes, respectively. We plot them based on their gender and education. Figure \ref{fig:slope-gender} shows top-5 slopes are dominated by males, while females dominate bottom-5 slopes. However, as this figure only shows a small portion of the data, we further inspect the gender wage gap.

```{r slope-gender, fig.cap = "Wages slope over the years based on gender. Most of respondents who experienced high wages slope are males. In contrast, respondents who experienced lower wages slope are mostly females."}
```

According to @OECD, the gender pay gap is the difference between the median wage of males and females relative to the median wage of males. We find that, as shown in \ref{fig:gap-gender}, over the years, females tend to earn fewer than males. For example, in 2018, females earned about 78 cents for every $1 earned by males. The widest gender gaps were observed in the 1990s. However, as more males participated in the survey than females from year to year, these gaps could be affected by imbalance number of observations.

```{r gap-gender, fig.cap="A) Number of observations by gender and by year. There are more males participated in survey than females. The number of observations decreased over the years. B) Gender wage gap or the median wages of females relatives to median wages of males. Females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every \\$1 earned by males. Note that this is a crude measure as gap is calculated not based on other variable, such as experience and job class.", out.width="90%"}
```

According to their highest education completed, both top-5 and bottom-5 slopes are dominated by people who completed high school. This is understandable since the respondents are mostly completed high school.

```{r slope-educ, fig.cap = "Wages slope over the years based on highest education completed. People who are in top-5 and bottom-5 wages are dominated by people who completed high school."}
```

Further, we perform a linear regression to see the effect of education on wages experience.

>>>>

## Recap

@@

# Exploratory Data Analysis
